Combinatorial interpretations of holonomic recurrences

### 2021-08-19 article/main.tex
Fix spelling

\(\tSub(v)\) is implicitly defined as a whole subree rooted with node \(v\).

\(\tLeft(v)\) is implicitly defined as a left subree of node \(v\).

\(\tRight(v)\) is implicitly defined as a right subree of node \(v\).

@@

In order to get some intuition on how it works unbeneath, let's take a look on how we can find number of the full binary trees of size \(2\). We can expand the recurrence a bit:

@@

\begin{example}
    Sequences \(a_n = \sqrt{n}\), \(b_n = n^n\) are \textbf{not} holonomic, ie. none of them satisifies a linear recurrence equation with polynomial coefficients\cite{nonholo}.
\end{example}

@@

Holonomic recurrences are very useful if we want to count objects of given size. Of course, size can be defined in more complex ways, eg.\ we may want to find number of trees with \(n\) internal nodes having exactly \(m\) nodes beging left child of its parent.

@@

It is easy to see that tree of size \(n\) (\(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblence to our holonomic recurrence, you are perfectly right.

@@

In the second case it is enough to noitice that \(R_l\) generates trees with pointed node being a left child, and \(R_r\) generates trees with pointed node being right child. That immediately provides us with conclusion that we are generating no duplicates.

@@

The definition above is precise, but introduces a new kind of vertice that we would have to take care of. We can avoid it by using a trick to make the tree recall binary tree. We don't change the structure, just the notation.

@@

When aplying \(L_2\) we achieve following cases respectively:

@@

To achive trees from \((n - 2) T(n - 1)\), we can simply take trees from case 3. and replace the father of the \(s\) with \(t''\) and point root of \(t''\). In this way, we will loose \(2\) leaves and obtain each tree from \((n - 2) T(n - 1)\).

@@

As you could see in the examples above, most of the solutions base on the simple analysis of number of laves, internal nodes and other similar properties. In this section, we've gathered some methods that may be helpful when trying to analyze the problem of interpreting combinatorically some holonomic recurrence.

@@

First thing that we have to select is defining the size carrefully. We have few standard choices of computing tree sizes:
\begin{itemize}
    \item counting all existing nodes,
    \item counting internal nodes,
    \item counting leaves,
    \item counting null-nodes (ie.\ children of the leaves),
    \item edges.
\end{itemize}

@@

As the main field of our interest was analyzis of terms of lambda trees in unary de Bruijn notation, we will use it as an example to describe the method.\todo{Add definition}

@@

\begin{definition}
    Set of lambda terms \(\Lambda\) is defined recursively:
    \begin{itemize}
        \item if \(x\) is a variable, \(x \in \Lambda\)
        \item if \(M, N \in \Lambda\), then \((M N) \in \Lambda\) (application of function \(M\) to an argument \(N\))
        \item if \(x\) is a variable and \(M \in \Lambda\), then \((\lambda x . M) \in \Lambda\) (abstraction, ie.\ creating a function with an argument \(x\) and binding each occurence of \(x\) in \(M\) to this argument)
    \end{itemize}
\end{definition}

@@

Which leads to following interpreation of \(\Phi\):

@@

\subsection{Enhanced holonomic reccurence}%
\label{sub:enhanced_holonomic_reccurence}

@@

The latest two are intruduced in order to make the analysis simpler, but do not have immediate interpretation in the world of holonomic recurrences. If we use them both, it sums up to all nodes except root (as root is neither left nor right child).

@@

# old visitior
c.append_r_class_subtree_visitor(-1,
    lambda kind, left, right:
        None if kind[0] is None else
        ["l", kind] + left + right + [(None, "*")]
)

# new visitior
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["l", parent_kind] + [left[0] + ("*",)] + left[1:] + right + [None]
)

# new visitior
c.append_r_class_left_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["l", parent_kind] + left + [right[0] + ("*",)] + right[1:] + [None]
)

@@

# old visitior
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", (None, "*"), parent_kind] + left + right
)

# new visitior
c.append_r_class_left_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", None, parent_kind] + left + [right[0] + ("*",)] + right[1:]
)

# new visitior
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", None, parent_kind] + [left[0] + ("*",)] + left[1:] + right
)
\end{lstlisting}

Basicaly, for each virtual node not being root (ie.\ having left or right parent), we point it and add \(l\) or \(s\) respectively above the parent.

@@

For each actual vertex, we add \(a\) node as its left or right parent and add as a second child \(o\) node as its siblig. We then point left or right virtual child of newly created \(o\) node.

@@

You may also note that we are operating on specific \(N\). The probability that we miss some corner case is low, but even if we check large \(N\) values, it is always present. The formal algebraic proof would make us sure that the construction works. We woulr prioritize finishing the construction first.

>>>>

\(\tSub(v)\) is implicitly defined as a whole subtree rooted with node \(v\).

\(\tLeft(v)\) is implicitly defined as a left subtree of node \(v\).

\(\tRight(v)\) is implicitly defined as a right subtree of node \(v\).

@@

In order to get some intuition on how it works underneath, let's take a look on how we can find number of the full binary trees of size \(2\). We can expand the recurrence a bit:

@@

\begin{example}
    Sequences \(a_n = \sqrt{n}\), \(b_n = n^n\) are \textbf{not} holonomic, ie. none of them satisfies a linear recurrence equation with polynomial coefficients\cite{nonholo}.
\end{example}

@@

Holonomic recurrences are very useful if we want to count objects of given size. Of course, size can be defined in more complex ways, eg.\ we may want to find number of trees with \(n\) internal nodes having exactly \(m\) nodes being left child of its parent.

@@

It is easy to see that tree of size \(n\) (\(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

@@

In the second case it is enough to notice that \(R_l\) generates trees with pointed node being a left child, and \(R_r\) generates trees with pointed node being right child. That immediately provides us with conclusion that we are generating no duplicates.

@@

The definition above is precise, but introduces a new kind of node that we would have to take care of. We can avoid it by using a trick to make the tree recall binary tree. We don't change the structure, just the notation.

@@

When applying \(L_2\) we achieve following cases respectively:

@@

To achieve trees from \((n - 2) T(n - 1)\), we can simply take trees from case 3. and replace the father of the \(s\) with \(t''\) and point root of \(t''\). In this way, we will loose \(2\) leaves and obtain each tree from \((n - 2) T(n - 1)\).

@@

As you could see in the examples above, most of the solutions base on the simple analysis of number of laves, internal nodes and other similar properties. In this section, we've gathered some methods that may be helpful when trying to analyze the problem of interpreting combinatorially some holonomic recurrence.

@@

First thing that we have to select is defining the size carefully. We have few standard choices of computing tree sizes:
\begin{itemize}
    \item counting all existing nodes,
    \item counting internal nodes,
    \item counting leaves,
    \item counting null-nodes (ie.\ children of the leaves),
    \item edges.
\end{itemize}

@@

As the main field of our interest was analysis of terms of lambda trees in unary de Bruijn notation, we will use it as an example to describe the method.\todo{Add definition}

@@

\begin{definition}
    Set of lambda terms \(\Lambda\) is defined recursively:
    \begin{itemize}
        \item if \(x\) is a variable, \(x \in \Lambda\)
        \item if \(M, N \in \Lambda\), then \((M N) \in \Lambda\) (application of function \(M\) to an argument \(N\))
        \item if \(x\) is a variable and \(M \in \Lambda\), then \((\lambda x . M) \in \Lambda\) (abstraction, ie.\ creating a function with an argument \(x\) and binding each occurrence of \(x\) in \(M\) to this argument)
    \end{itemize}
\end{definition}

@@

Which leads to following interpretation of \(\Phi\):

@@

\subsection{Enhanced holonomic recurrence}%
\label{sub:enhanced_holonomic_reccurence}

@@

The latest two are introduced in order to make the analysis simpler, but do not have immediate interpretation in the world of holonomic recurrences. If we use them both, it sums up to all nodes except root (as root is neither left nor right child).

@@

# old visitor
c.append_r_class_subtree_visitor(-1,
    lambda kind, left, right:
        None if kind[0] is None else
        ["l", kind] + left + right + [(None, "*")]
)

# new visitor
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["l", parent_kind] + [left[0] + ("*",)] + left[1:] + right + [None]
)

# new visitor
c.append_r_class_left_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["l", parent_kind] + left + [right[0] + ("*",)] + right[1:] + [None]
)

@@

# old visitor
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", (None, "*"), parent_kind] + left + right
)

# new visitor
c.append_r_class_left_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", None, parent_kind] + left + [right[0] + ("*",)] + right[1:]
)

# new visitor
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", None, parent_kind] + [left[0] + ("*",)] + left[1:] + right
)
\end{lstlisting}

Basically, for each virtual node not being root (ie.\ having left or right parent), we point it and add \(l\) or \(s\) respectively above the parent.

@@

For each actual vertex, we add \(a\) node as its left or right parent and add as a second child \(o\) node as its sibling. We then point left or right virtual child of newly created \(o\) node.

@@

You may also note that we are operating on specific \(N\). The probability that we miss some corner case is low, but even if we check large \(N\) values, it is always present. The formal algebraic proof would make us sure that the construction works. We would prioritize finishing the construction first.
